---
title: "tcr-seek: TCR repertoire QC report"
subtitle: "Diversity and sample-level QC"
author:
  - tcr-seek
title-block-banner: "#27445C"
title-block-logo: "tcr-seek.svg"
format:
  html:
    lightbox: true
    toc: true
    toc-location: left
    page-layout: full
    theme:
      light: cosmo
      dark: darkly
    html-math-method: katex
    code-fold: true
    highlight-style: tango
    toc_float: true
    number-sections: false
    embed-resources: true
    smooth-scroll: true
    fig-align: "center"
    fig-width: 10
    fig-height: 8
execute:
  cache: false
editor: source
editor_options:
  chunk_output_type: console
params:
  immdata_rds: "03_immunarch/Immdata.rds"
  diversity_rds: "03_immunarch/diversity_summary.rds"
  qc_variables: ""
  logo: "resources/tcr-seek.svg"
---

Diversity metrics were calculated from the immunarch object using the `vegan` package in R. The default clonotype mode is `strict`, unless the pipeline was run with a different `--clonotype-definition`. Supported presets are `gene` (V/D/J/C genes when present), `nt` (CDR3 nucleotide), `aa` (CDR3 amino acid), and `strict` (CDR3 nucleotide plus V/D/J/C genes when present). A custom comma-separated immunarch column list can also be supplied.

- Richness is the number of observed clonotypes in a sample. It is intuitive, but it is highly sensitive to sequencing depth.
- Shannon diversity combines richness and evenness. It decreases when a small number of expanded clonotypes dominate the repertoire.
- vegan Simpson diversity is reported as `1 - D`, where `D = sum(p_i^2)`. Larger values indicate higher diversity and less dominance by abundant clonotypes.
- Evenness is Pielou evenness: Shannon diversity divided by `log(richness)`. It describes how evenly clone abundance is distributed across observed clonotypes.
- Library size is the total `Clones` abundance used in the diversity matrix.
- `RichnessPerMillion` is retained only as a rough descriptive normalization. Richness is nonlinear with depth, so dividing by reads-per-million can over-correct shallow samples and under-correct deep samples. Prefer depth-aware models for comparisons.

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, cache = FALSE, fig.align = "center", fig.width = 10, fig.height = 8)

suppressPackageStartupMessages({
  library(tidyverse)
  library(DT)
  library(ggrepel)
  library(plotly)
})

read_qc_variables <- function(value, metadata) {
  requested <- trimws(strsplit(value, ",", fixed = TRUE)[[1]])
  requested <- requested[nzchar(requested)]
  if (!length(requested)) {
    requested <- setdiff(colnames(metadata), "Sample")
  } else {
    available <- colnames(metadata)
    requested <- vapply(requested, function(variable) {
      match <- available[tolower(available) == tolower(variable)]
      if (length(match)) match[[1]] else NA_character_
    }, character(1))
    requested <- stats::na.omit(requested)
  }
  unique(intersect(requested, colnames(metadata)))
}

normalize_delivery_name <- function(delivery) {
  dplyr::case_when(
    delivery == "20260710_SH01446_0005_ASC2252557-SC3_From412" ~ "batch5",
    TRUE ~ delivery
  )
}

summarise_fastq_delivery <- function(delivery_table) {
  delivery_table %>%
    mutate(
      fastq.delivery = normalize_delivery_name(fastq.delivery),
      Sample = toupper(Sample)
    ) %>%
    filter(nzchar(Sample), nzchar(fastq.delivery)) %>%
    distinct(Sample, fastq.delivery) %>%
    group_by(Sample) %>%
    summarise(fastq.delivery = paste(sort(unique(fastq.delivery)), collapse = ";"), .groups = "drop")
}

read_fastq_delivery <- function(run_dir) {
  sample_fastq <- file.path(run_dir, "sample_fastq.tsv")
  if (file.exists(sample_fastq)) {
    manifest_delivery <- readr::read_tsv(sample_fastq, show_col_types = FALSE, col_types = readr::cols(.default = readr::col_character())) %>%
      transmute(
        Sample = sample,
        fastq.delivery = basename(dirname(fastq))
      ) %>%
      summarise_fastq_delivery()
    if (nrow(manifest_delivery)) {
      return(manifest_delivery)
    }
  }

  source_dir <- file.path(run_dir, "input_fastqs")
  source_files <- list.files(source_dir, pattern = "\\.R[12]\\.fastq\\.gz\\.sources\\.txt$", full.names = TRUE)
  if (!length(source_files)) {
    return(tibble(Sample = character(), fastq.delivery = character()))
  }

  purrr::map_dfr(source_files, function(source_file) {
    sample <- sub("\\.R[12]\\.fastq\\.gz\\.sources\\.txt$", "", basename(source_file))
    source_paths <- readLines(source_file, warn = FALSE)
    delivery_dirs <- source_paths %>%
      trimws() %>%
      discard(~ !nzchar(.x)) %>%
      dirname() %>%
      basename() %>%
      unique() %>%
      sort()
    tibble(Sample = sample, fastq.delivery = delivery_dirs)
  }) %>%
    summarise_fastq_delivery()
}

metric_cols <- c("Richness", "Shannon", "Simpson", "Evenness", "LibrarySize", "RichnessPerMillion", "EffectiveClonotypes", "TopCloneFraction", "Top10CloneFraction")
immdata <- readRDS(params$immdata_rds)
diversity <- readRDS(params$diversity_rds)
definition_comparison_rds <- file.path(dirname(params$diversity_rds), "diversity_by_clonotype_definition.rds")
definition_comparison <- if (file.exists(definition_comparison_rds)) readRDS(definition_comparison_rds) else diversity
metadata <- as.data.frame(immdata$meta) %>% distinct(Sample, .keep_all = TRUE)
run_dir <- dirname(dirname(normalizePath(params$immdata_rds, mustWork = TRUE)))
metadata <- metadata %>% left_join(read_fastq_delivery(run_dir), by = "Sample")
qc_variables <- read_qc_variables(params$qc_variables, metadata)
metadata <- metadata %>%
  mutate(across(all_of(qc_variables), ~ factor(dplyr::na_if(as.character(.x), ""))))

diversity <- diversity %>% left_join(metadata, by = "Sample")
definition_comparison <- definition_comparison %>% left_join(metadata, by = "Sample")
long_diversity <- diversity %>%
  pivot_longer(cols = all_of(intersect(metric_cols, colnames(diversity))), names_to = "Metric", values_to = "Value")
long_definition_comparison <- definition_comparison %>%
  pivot_longer(cols = all_of(intersect(metric_cols, colnames(definition_comparison))), names_to = "Metric", values_to = "Value")

create_dt <- function(table, download_basename = "tcr_seek_qc") {
  DT::datatable(
    table,
    rownames = FALSE,
    filter = "top",
    extensions = c("Buttons", "Scroller", "SearchBuilder", "FixedColumns"),
    width = "100%",
    options = list(
      dom = "QBlrtip",
      scrollX = TRUE,
      scrollY = 500,
      scroller = TRUE,
      fixedColumns = list(leftColumns = 1),
      buttons = list(
        list(extend = "csv", filename = download_basename),
        list(extend = "excel", filename = download_basename)
      )
    )
  )
}
```

## Selected Definition Summary

```{r clonotype-definition-summary, results="asis"}
selected_definition <- diversity %>%
  distinct(ClonotypeMode, ClonotypeDefinition)
selected_mode <- as.character(unlist(selected_definition$ClonotypeMode))[1]
selected_columns <- as.character(unlist(selected_definition$ClonotypeDefinition))[1]
cat("**Selected clonotype mode:** `", selected_mode, "`  \n", sep = "")
cat("**Resolved clonotype columns:** `", selected_columns, "`  \n", sep = "")
cat("**Rarefaction:** not run for this report.\n\n", sep = "")
```

```{r}
summary_table <- definition_comparison %>%
  arrange(Sample, factor(ClonotypeMode, levels = c("gene", "nt", "aa", "strict", "custom"))) %>%
  mutate(
    ClonotypeMode = factor(ClonotypeMode, levels = c("gene", "nt", "aa", "strict", "custom")),
    ClonotypeDefinition = factor(ClonotypeDefinition, levels = unique(ClonotypeDefinition))
  ) %>%
  select(any_of(c("Sample", "ClonotypeMode", "ClonotypeDefinition", metric_cols, qc_variables)))
create_dt(summary_table, "tcr_seek_diversity_summary")
```

## Clonotype Definition Comparison

```{r}
comparison_metrics <- c("Richness", "Shannon", "Simpson", "Evenness", "EffectiveClonotypes", "TopCloneFraction", "Top10CloneFraction")
long_definition_comparison %>%
  filter(Metric %in% comparison_metrics) %>%
  ggplot(aes(x = ClonotypeMode, y = Value, color = ClonotypeMode)) +
  geom_point(position = position_jitter(width = 0.12, height = 0), alpha = 0.7, size = 1.8) +
  geom_line(aes(group = Sample), color = "grey70", alpha = 0.45) +
  facet_wrap(~Metric, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 35, hjust = 1), legend.position = "none") +
  labs(x = "Clonotype definition mode", y = "Value")
```

```{r}
long_diversity %>%
  ggplot(aes(x = Metric, y = Value)) +
  geom_boxplot(outlier.shape = NA, fill = "grey85") +
  geom_jitter(width = 0.2, alpha = 0.45, size = 1.4) +
  facet_wrap(~Metric, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(x = NULL, y = "Value")
```

## QC Variables

```{r qc-variable-note, results="asis"}
if (!length(qc_variables)) {
  cat("No QC variables were supplied or found in sample metadata. Add `--sample-info` and `--qc-variables` to group diversity metrics by sample annotations.\n")
}
```

```{r qc-variable-tabs, results="asis"}
if (length(qc_variables)) {
  definition_levels <- c("gene", "nt", "aa", "strict", "custom")
  available_definitions <- intersect(definition_levels, unique(as.character(definition_comparison$ClonotypeMode)))
  cat("::: {.panel-tabset}

")
  for (variable in qc_variables) {
    cat("### ", variable, "

", sep = "")
    cat("::: {.panel-tabset}

")
    for (definition_mode in available_definitions) {
      definition_label <- definition_comparison %>%
        filter(ClonotypeMode == definition_mode) %>%
        distinct(ClonotypeDefinition) %>%
        pull(ClonotypeDefinition) %>%
        as.character() %>%
        .[1]
      cat("#### ", definition_mode, "

", sep = "")
      cat("`", definition_label, "`

", sep = "")

      plot_data <- long_definition_comparison %>%
        filter(ClonotypeMode == definition_mode, !is.na(.data[[variable]]), .data[[variable]] != "")
      if (!nrow(plot_data)) {
        cat("No non-missing values available for this variable and clonotype definition.

")
        next
      }
      print(
        ggplot(plot_data, aes(x = .data[[variable]], y = Value, fill = .data[[variable]])) +
          geom_boxplot(outlier.shape = NA, alpha = 0.75) +
          geom_jitter(width = 0.18, alpha = 0.45, size = 1.3) +
          facet_wrap(~Metric, scales = "free_y") +
          theme_bw() +
          theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") +
          labs(x = variable, y = "Value")
      )
      cat("

")

      numeric_metrics <- definition_comparison %>%
        filter(ClonotypeMode == definition_mode) %>%
        select(any_of(metric_cols))
      if (ncol(numeric_metrics)) {
        tab <- definition_comparison %>%
          filter(ClonotypeMode == definition_mode) %>%
          group_by(.data[[variable]]) %>%
          summarise(across(all_of(colnames(numeric_metrics)), list(median = ~median(.x, na.rm = TRUE), mean = ~mean(.x, na.rm = TRUE)), .names = "{.col}_{.fn}"), n = n(), .groups = "drop") %>%
          rename(!!variable := 1)
        print(create_dt(tab, paste0("tcr_seek_qc_", variable, "_", definition_mode)))
        cat("

")
      }
    }
    cat(":::

")
  }
  cat(":::
")
}
```
